Fitting Multifactorial Models of Differential Expression

### 2022-07-19 DGE_comparison_v2.Rmd
Your Message

```{r,message = FALSE,echo = FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
opts_chunk$set(fig.width = 10,
               fig.height = 5,
               cache = FALSE)
```
---
title: "Fitting Multifactorial Models of Differential Expression"
author: "Samuel N Bogan"
bibliography: DGE_Comp.bib
---

@@

|           Studies of molecular responses to environmental change increasingly employ multifactorial experimental designs incorporating different developmental stages, stressors, populations, or non-linear dynamics that resolve interactions between variables that shape expression, physiology, and performance. After an experiment during which a study organism was exposed to abiotic variation and perhaps other variables, researchers interested in how gene expression correlates with these variables will often extract RNA across treatment groups, prepare cDNA libraries for sequencing, and conduct an analysis of differential expression (DE) across variables of interest after mapping and counting RNA-seq reads. DE analysis is achieved by fitting RNA-seq read count data to statistical models, often using tried-and-true DE packages. Most packages do not have the flexibility to accommodate a number of common experimental design components and it can be difficult to navigate their functionality for fitting multivariate models. When these packages fall short, determining how to fit customized multifactorial models yourself can be an additional challenge. Many great walkthroughs exist for conducting DE analysis across using single-variable experimental designs, but there is a lack of guidance on multifactorial approaches for testing for DE and modelling gene expression. Regardless of whether you have RNA-seq data on hand or you're in the process of planning your experiment, this walkthrough will help you navigate what DE packages and model fitting approaches are best suited to your experimental design.

## Targeted audience  
|           This walkthrough is for both intermediate and expert practitioners of RNA-seq analyses, and most of all, those new to multifactorial RNA-seq. We have written this page with the assumption that readers have a basic familiarity with model fitting and statistical analysis, as well as next generation sequencing technologies, RNA-seq workflows, and the R statistical environment.

@@

|           This walkthrough can be broken down into different types of predictor variables that can be incorporated in models of gene expression. In the order we address them, these variables include:

@@

@@

|           Two important considerations when choosing a DE package to analyze RNA-seq data are the distribution family used by the package and the effect types it incorporate into a model. The DE packages we will cover assume the following read count distributions and are capable of fitting the following effect types:

| Program | Distribution | Continuous fixed eff. | Random intercepts | Random slopes | Interactive eff. | Non-linear eff. |
| :--- | :--- | :--- | :--- | :--- | :--- |
| edgeR | Negative binomial generalized linear model | ✔ | ✖ |  ✖ |  ✔ |  ✔ |  
| DESeq2 | Negative binomial generalized linear model | ✔ | ✖ | ✖ | ✔|   ✔ | 
| Voom | Mean-variance linear model | ✔ | ✔ | ✖ |  ✔|   ✔ |

@@

|           We will walk you through (i) how to conduct essential checks of your data such as principal components analysis and observing the distribution of filtered read counts, (ii) evaluating each package's functionality for multifactorial model fitting, and (iii) how to fit each type of predictor by performing differential expression tests using real data. These data come from an experiment that exposed the Antarctic pteropod (a free-swimming planktonic snail) *Limacina helicina antarctica* to three continuous levels of ocean acidification (OA) for either 0.5 or 7 days [@Bogan2020-qs]. These RNA-seq counts are derived from reads mapped and counted via RSEM [@Li2011-mi] using a *de novo* reference transcriptome that includes transcript isoforms assembled by @Johnson2016-wq.

|           Throughout this walkthrough, we will provide examples of custom scripts for fitting linear, linear mixed models, generalized linear models to read count data using for loops in order to provide functionality where it cannot be met by most differential expression packages.

|           At the end of the walkthrough, we demonstrate how to compare model predictions by different packages for different effect types commonly used in multifactorial RNAseq analyses.

|           Let's start by loading our required packages...

```{r setup, include = FALSE}
# should remove from Rmd
# Set root directory
#knitr::opts_knit$set( root.dir = '~/Documents/GitHub/DGE_comparison/' )

@@

|           Before model fitting and testing for DE, it is important to visually inspect read counts pre- and post-filtering. In this walk through, we will plot the distributions of read counts and sample loading to multidimensional scaling axes (e.g., plotting principal components or principal coordinates analyses). Looking at these plots, we can (i) determine whether the distribution of our data match the assumptions of model families used by different packages and (ii) determine whether one or multiple variables within a dataset explain a significant degree of variation in gene expression across replicates. Depending on the distribution of your data and the variables that appear important (or variable you would like to incorporate into your models), we will then demonstrate how to choose an appropriate package to test for DE.
|           Most DE packages assume that read counts possess a negative binomial distribution. The negative binomial distribution is an extension of distributions for binary variables such as the Poisson distribution, allowing for estimations of "equidispersion" and "overdispersion", equal and greater-than-expected variation in expression attributed to biological variability . However, RNA-seq datasets can exhibit poor fits with the negative binomial distribution [@Hawinkel2020-lk]. It is well worth visualizing and testing the distributions of RNA-seq read counts before and after filtering. Keep in mind that the distribution applies not to the raw data, but the residuals of a model. Additionally, individual models are fit to the read counts of each gene. Thus, the most accurate representation of how appropriate a negative binomial distribution is for read counts would come from plotting the distribution of residuals across all genes following model fitting. Below is a visual example of negative binomial distributions that possess different variances from modelassist[dot]epixanalytics[dot]com:

```{r, echo = FALSE, out.height = "300%", fig.align = "center", }

knitr::include_graphics("images/negbinom.png")

@@

|           Now let's first plot the distribution of unfiltered reads from our dataset:

@@

|           The table of read counts above has rows representing individual transcripts and columns representing different replicate samples across treatments.

```{r, message = FALSE, warning = FALSE}
# Name experimental samples: names correspond to pCO2 treatment (300, 600, 900) + days (12 h2 vs 7 days) + replicate ID
colnames(data) <- c( "300.7.a", "300.7.b", "300.7.c","300.12.a", "300.12.b", "300.12.c", 
                       "900.7.a", "900.7.b", "900.7.c", "900.12.a", "900.12.b", "900.12.c", 
                       "600.7.a", "600.7.b", "600.7.c", "600.12.a", "600.12.b", "600.12.c")

## Create 'targets' and 'Group dataframe, expressing experimental variables for DEG analysis
pCO2 <- as.numeric( c( 255, 255, 255, 255, 255, 255,
                       530, 530, 530, 530, 530, 530,
                       918, 918, 918, 918, 918, 918 ) )

treatment <- c( "B", "B", "B", "B", "B", "B",
                "R", "R", "R", "R", "R", "R",
                "Y", "Y", "Y", "Y", "Y", "Y" )

day <- as.numeric( c( 7, 7, 7, .5, .5, .5,
                      7, 7, 7, .5, .5, .5,
                      7, 7, 7, .5, .5, .5 ) )

targets <- data.frame( pCO2, day, treatment )
targets$grouping <- paste( targets$pCO2, targets$day, sep = "." )

# The group factor represents the combined levels of an experimental replicate across all variables
Group <- factor( paste( targets$day, targets$pCO2, sep = "_" ) )

# Data must be rounded to nearest integer in order to be fit for negative binomial distribution
data_input <- round( data )

# Peak at rounded data
head( data_input )

# Plot distribution of unfiltered read counts across all samples 
ggplot( data = data.frame( rowMeans( data_input ) ),
        aes( x = rowMeans.data_input. ) ) +
  geom_histogram( fill = "grey" ) +
  xlim( 0, 500 ) +
  theme_classic() +
  labs( title = "Distribution of unfiltered reads" ) +
  labs( y = "Density", x = "Raw read counts", title = "Read count distribution: untransformed, unnormalized, unfiltered" )

@@

|           As you can see in the above plot, the raw distribution of all read counts takes on a left-skewed negative binomial distribution similar to the purple distribution in the example above. While looking at the distribution of your raw reads is useful, these are not in fact the data you will be inputting to tests of differential expression. Let's plot the distribution of filtered reads normalized by library size, expressed as log2 counts per million reads (logCPM). These are the reads we will use in our test, and after we plot their distribution, we will conduct one more, slightly more robust, test of our data's fit to the negative binomial distribution using residuals from fitted models.

@@

# Make a DGEList object for edgeR
y <- DGEList( counts = data_input, remove.zeros = TRUE )

#Let's remove samples with less than 0.5 cpm (this is ~10 counts in the count file) in fewer then 9/12 samples
keep <- rowSums( cpm( y ) > .5 ) >= 9

table( keep )

# Set keep.lib.sizes = F and recalculate library sizes after filtering
y <- y[ keep, keep.lib.sizes = FALSE ]

y <- calcNormFactors( y )

# Calculate logCPM
df_log <- cpm( y, log = TRUE, prior.count = 2 )

# Plot distribution of filtered logCPM values
ggplot( data = data.frame( rowMeans( df_log ) ),
        aes( x = rowMeans.df_log. ) ) +
  geom_histogram( fill = "grey" ) +
  theme_classic() +
  labs( y = "Density", x = "Filtered read counts (logCPM)", 
        title = "Distribution of normalized, filtered read counts" )

@@

|           Our raw data appear to follow a strong negative binomial distribution! Later, we will see whether this holds for residuals from fitted negative binomial GLMs.

@@

|           Before analyzing our data, it is essential that we look at the multivariate relationships between our samples based on transcriptome-wide expression levels. Below is example code and output for a principal coordinates analysis (PCOA) plot that visualizes multifactorial RNA-seq replicates according to two predictor variables across major and minor latent variables or PCOA axes. These predictor variables, as discussed above, are *p*CO${_2}$ and time-in-treatment.

@@

# Export pcoa loadings
dds.pcoa = pcoa( vegdist( t( df_log <- cpm( y, log = TRUE, prior.count = 2 ) ),
                          method = "euclidean") / 1000 )

@@

# Calculate % variation explained by each eigenvector
percent <- dds.pcoa$values$Eigenvalues
cumulative_percent_variance <- ( percent / sum( percent ) ) * 100

# Prepare information for pcoa plot, then plot
color <- c( "steelblue1", "tomato1", "goldenrod1")
par( mfrow = c( 1, 1 ) )
plot( scores[ , 1 ], scores[ , 2 ],
      cex=.5, cex.axis=1, cex.lab = 1.25,
      xlab = paste( "PC1, ", round( cumulative_percent_variance[ 1 ], 2 ), "%" ), 
      ylab = paste( "PC2, ", round( cumulative_percent_variance[ 2 ], 2 ), "%" ) )

# Add visual groupings to pcoa plot
ordihull( scores, as.factor(targets$treatment ), 
          border = NULL, lty = 2, lwd = .5, label = F,
          col = color, draw = "polygon", alpha = 100, cex = .5 )

ordispider( scores,as.factor( targets$grouping ),label = F ) # Vectors connecting samples in same pCO2 x time group

ordilabel( scores, cex = 0.5) # Label sample IDs

@@

|           In the PCOA above, blue = low *p*$CO_{2}$, yellow = moderate *p*$CO_{2}$, and red = high *p*$CO_{2}$. From this plot, we can see that treatment and time both influence multivariate gene expression across the RNA-seq samples. By and large, samples cluster according to these two predictors in a manner consistent with what we would expect from our experimental design if *p*$CO_{2}$ and time were to affect gene expression.

@@

logCPM.pca <- prcomp( t ( df_log ) )
logCPM.pca.proportionvariances <- ( ( logCPM.pca$sdev ^ 2 ) / ( sum( logCPM.pca$sdev ^ 2 ) ) ) * 100

## Do treatment groups fully segregate? Wrap samples by pCO2 x time, not just pCO2
# Replot using logCPM.pca
plot( logCPM.pca$x, type = "n", main = NA, xlab = paste( "PC1, ", round( logCPM.pca.proportionvariances[ 1 ], 2 ), "%" ), ylab = paste( "PC2, ", round( logCPM.pca.proportionvariances[ 2 ], 2 ), "%" ) )

points( logCPM.pca$x, col = "black", pch = 16, cex = 1 )
colors2 <- c( "steelblue1", "dodgerblue2", "tomato1", "coral", "goldenrod1", "goldenrod3" )

ordihull( logCPM.pca$x, targets$grouping, 
          border = NULL, lty = 2, lwd = .5, 
          col = colors2, draw = "polygon", 
          alpha = 75,cex = .5, label = T )

@@

|           This plot is only a slight adjustment of the first PCOA. If you'd like to wrap each multifactorial group and get a better idea of potential overlap between clusters, the plot and code above can be applied to do so.

|           Now that we have conducted QC analyses and plots of the read count data, it is time to begin model fitting and testing for differential expression.

@@

|           Gene expression traits can exhibit non-linear performance curves across a continuous variable much like other physiological traits. Studies that incorporate more than two levels of a continuous predictor provide an opportunity to test for non-linear effects on gene expression. Similar to performance traits such as growth or metabolic rates, gene expression can be modelled across a continuous variable by fitting non-linear 'performance curves' to read count data. Rivera *et al.* 2021 recently discussed the benefits of fitting non-linear reaction norms to gene expression data in order to better model acclimation by tolerant and sensitive genotypes to an environmental stressor. In a conceptual figure of theirs, shown below, they depict non-linear variation in expression before, during, and after exposure to a stressor for four reaction norm phenotypes: (i) a tolerant conspecific that frontloads (e.g., constitutively upregulates) stress-induced transcripts, (ii) a tolerant conspecific that exhibits reduced overall expression of stress-responsive transcripts, (iii) a sensitive conspecific that upregulates inducible transcripts and cannot recover post-stress, and (iv) a resilient conspecific that mounts a transcriptional stress response followed by recovery and downregulation [@Rivera2021-hn].

```{r, echo = FALSE, out.height = "40%", fig.align = "center", }
knitr::include_graphics("images/Rivera_etal_fig.png")
```
From Rivera *et al.* 2021 - "Transcriptomic resilience and timing. (a) Gene expression reaction norms of four strategies during recovery after a stressor. We use triangles again for patterns that may confer tolerance and circles for patterns associated with stress sensitivity. While all triangle paths show a return to baseline (resilience) the pink (frontloading) and yellow (dampening) are also depicting differences in baseline and plasticity and are therefore labelled differently. (b) Adapted from the rolling ball analogy commonly used for ecological resilience and depicted in Hodgson et al. (2015). Each ball represents a gene showing a color-matched expression pattern in (a). Landscapes represent expression possibilities during a stress event. In the absence of stress, the ball will settle in a trough, representing baseline expression levels. Elasticity (rate of return to the baseline) is represented by the size of the arrow (i.e., larger arrows have faster rates of return). Pink dotted line is the expression landscape for the frontloaded ball. (c) Using Torres et al. (2016) loops through disease space as an alternative framework of an organism's path through stress response and recovery. The color gradient represents the resulting phenotype for a given path through stress and recovery space, though x-and y-axis can denote any two parameters that are correlated but with a time lag."

|           Non-linear reaction norms can be modelled across time, such as Rivera *et al.* have presented, or across multiple levels of an abiotic variable such as temperature, pH, etc. in order to fit a performance curve. Performance curves are a fundamental tool in ecological physiology, and enable more robust hypothesis testing in RNA-seq studies of environmental acclimation. For example, testing for variation in the shape of gene expression performance curves between early and long-term exposure timepoints can provide critical information about the role of gene expression in acclimation. Below we outline how to fit and test for non-linear gene expression performance curves in multifactorial RNA-seq experiments using examples in edgeR, DESeq2, and custom code.    
|           One of the simplest non-linear relationships that can be fitted to the expression of a transcript across an continuous variable is a second-order polynomial, otherwise known as a quadratic function, which can be expressed as $y_{i} = \mu + \beta_{1}x^2 + \beta_{2}x$ where $y$ is the abundance of a given transcript ($i$), $\mu$ is the intercept, and $y$ is the continuous variable. For the parabola generated by fitting a second-order polynomial, $\beta_{1}$ > 0 opens the parabola upwards while $\beta_{1}$ < 0 opens the parabola downwards. The vertex of the parabola is controlled by $\beta_{2}$ such that when $\beta_{1}$ is negative, greater $\beta_{2}$ values result in the vertex falling at higher values of $x$. 
|           Quadratic polynomials applied to phenotypic performance curves commonly possess negative $\beta_{1}$ values with positive $\beta_{2}$ values: a downard-opening parabola with a positive vertex. However, quadtratic curves fitted to gene expression data can take on a variety of postiive or negative forms similar to exponential curves, saturating curves, and parabolas. For instance, the expression of a gene may peak an intermediate level of an environmental level before crashing or it may exponentially decline across that variable. Such trends may better model changes in the transcription of a gene compared to a linear model. To get started, we will fit non-linear second order polynomials before testing for whether model predictions for a given gene are significantly improved by a non-linear linear model.

@@

|           Let's fit a second-order polynomial for the effect of *p*CO$_{2}$ using edgeR. Using differential expression tests, we will then determine whether *p*CO$_{2}$ affected a gene's rate of change in expression and expression maximum by applying differential expression tests to $\beta_{1}$ and $\beta_{2}$ parameters. By testing for differential expression attributed to intereactions between time and $\beta_{1}$ or $\beta_{2}$, we will then test for whether these parameters were significantly different across exposure times such that 0.5 days and 7 days of acclimation to *p*CO$_{2}$ altered the rate of change in expression across *p*CO$_{2}$ ($\beta_{1}$) or the maximum of expression ($\beta_{2}$).

|          Since this is the first time we're fitting models to our data in this walkthrough, we will also output some important plots to help us explore and QC model predictions as we work toward testing for differential expression attributed to non-linear effects.

@@

# Square pCO2 variable
pCO2_2 <- pCO2^2

# Estimate dispersion coefficients
y1 <- estimateDisp( y, robust = TRUE ) # Estimate mean dispersal

# Plot tagwise dispersal and impose w/ mean dispersal and trendline
plotBCV( y1 )

@@

|           Our residuals appear to be normally distributed, indicating that our data fit the negative binomial distribution assumed by the GLM.

|           Now let's get back to analyzing non-linear effects on gene expression!

@@

# Filter tabularized logCPM df for read depth of logCPM >= 0.5 in > 75% of samples

>>>>

---
title: "Fitting Multifactorial Models of Differential Expression"
subtitle: '*Samuel N. Bogan*'
bibliography: DGE_Comp.bib
---

@@

Studies of molecular responses to environmental change increasingly employ multifactorial experimental designs incorporating different developmental stages, stressors, populations, or non-linear dynamics that resolve interactions between variables that shape expression, physiology, and performance. After an experiment during which a study organism was exposed to abiotic variation and perhaps other variables, researchers interested in how gene expression correlates with these variables will often extract RNA across treatment groups, prepare cDNA libraries for sequencing, and conduct an analysis of differential expression (DE) across variables of interest after mapping and counting RNA-seq reads. DE analysis is achieved by fitting RNA-seq read count data to statistical models, often using tried-and-true DE packages. Most packages do not have the flexibility to accommodate a number of common experimental design components and it can be difficult to navigate their functionality for fitting multivariate models. When these packages fall short, determining how to fit customized multifactorial models yourself can be an additional challenge. Many great walkthroughs exist for conducting DE analysis across using single-variable experimental designs, but there is a lack of guidance on multifactorial approaches for testing for DE and modelling gene expression. Regardless of whether you have RNA-seq data on hand or you're in the process of planning your experiment, this walkthrough will help you navigate what DE packages and model fitting approaches are best suited to your experimental design.

<br>

## Targeted audience  
This walkthrough is for both intermediate and expert practitioners of RNA-seq analyses, and most of all, those new to multifactorial RNA-seq. We have written this page with the assumption that readers have a basic familiarity with model fitting and statistical analysis, as well as next generation sequencing technologies, RNA-seq workflows, and the R statistical environment.

<br>

@@

This walkthrough can be broken down into different types of predictor variables that can be incorporated in models of gene expression. In the order we address them, these variables include:

@@

<br>

@@

Two important considerations when choosing a DE package to analyze RNA-seq data are the distribution family used by the package and the effect types it incorporate into a model. The DE packages we will cover assume the following read count distributions and are capable of fitting the following effect types:

| Program | Distribution | Continuous fixed eff. | Random intercepts | Random slopes | Interactive eff. | Non-linear eff. |
| :--- | :--- | :--- | :--- | :--- | :--- | :--- |
| edgeR | Negative binomial generalized linear model | ✔ | ✖ |  ✖ |  ✔ |  ✔ |  
| DESeq2 | Negative binomial generalized linear model | ✔ | ✖ | ✖ | ✔|   ✔ | 
| Voom | Mean-variance linear model | ✔ | ✔ | ✖ |  ✔|   ✔ |

<br>

@@

We will walk you through (i) how to conduct essential checks of your data such as principal components analysis and observing the distribution of filtered read counts, (ii) evaluating each package's functionality for multifactorial model fitting, and (iii) how to fit each type of predictor by performing differential expression tests using real data. These data come from an experiment that exposed the Antarctic pteropod (a free-swimming planktonic snail) *Limacina helicina antarctica* to three continuous levels of ocean acidification (OA) for either 0.5 or 7 days [@Bogan2020-qs]. These RNA-seq counts are derived from reads mapped and counted via RSEM [@Li2011-mi] using a *de novo* reference transcriptome that includes transcript isoforms assembled by @Johnson2016-wq.

Throughout this walkthrough, we will provide examples of custom scripts for fitting linear, linear mixed models, generalized linear models to read count data using for loops in order to provide functionality where it cannot be met by most differential expression packages.

At the end of the walkthrough, we demonstrate how to compare model predictions by different packages for different effect types commonly used in multifactorial RNAseq analyses.

Let's start by loading our required packages...

<br>

```{r setup, include = FALSE}
# should remove from Rmd
# Set root directory
#knitr::opts_knit$set(root.dir = '~/Documents/GitHub/DGE_comparison/')

@@

Before model fitting and testing for DE, it is important to visually inspect read counts pre- and post-filtering. In this walk through, we will plot the distributions of read counts and sample loading to multidimensional scaling axes (e.g., plotting principal components or principal coordinates analyses). Looking at these plots, we can (i) determine whether the distribution of our data match the assumptions of model families used by different packages and (ii) determine whether one or multiple variables within a dataset explain a significant degree of variation in gene expression across replicates. Depending on the distribution of your data and the variables that appear important (or variable you would like to incorporate into your models), we will then demonstrate how to choose an appropriate package to test for DE.

Most DE packages assume that read counts possess a negative binomial distribution. The negative binomial distribution is an extension of distributions for binary variables such as the Poisson distribution, allowing for estimations of "equidispersion" and "overdispersion", equal and greater-than-expected variation in expression attributed to biological variability . However, RNA-seq datasets can exhibit poor fits with the negative binomial distribution [@Hawinkel2020-lk]. It is well worth visualizing and testing the distributions of RNA-seq read counts before and after filtering. Keep in mind that the distribution applies not to the raw data, but the residuals of a model. Additionally, individual models are fit to the read counts of each gene. Thus, the most accurate representation of how appropriate a negative binomial distribution is for read counts would come from plotting the distribution of residuals across all genes following model fitting. Below is a visual example of negative binomial distributions that possess different means and variances:

```{r, echo = FALSE, message = FALSE, warning = FALSE}

## Simulate two negative binomial distributions of different variances
# First, great function for doing so
generate_distribution <- function(n, mus, sizes){
  tibble(n = n, 
       t_mu = mus, 
       t_size = sizes)%>% 
  mutate(rt = pmap(list(n, t_mu, t_size),
                     ~rnbinom(n = ..1, mu = ..2, size = ..3))) %>% 
  unnest(rt) 
}

set.seed(123)

# Second, simulate distributions with chosen parameters
rt_df <- 
  generate_distribution(n = 2000,
                        mus = 25,
                        sizes = .5)

rt_df <- rbind(rt_df,
      generate_distribution(n = 2000,
                        mus = 75,
                        sizes = 15))

# Plot distributions with histogram
ggplot(data = rt_df, aes(x = rt, fill = as.factor(t_size), group = as.factor(t_size))) +
  geom_histogram(data = filter(rt_df, t_mu == 25), alpha = 0.75) +
  geom_histogram(data = filter(rt_df, t_mu == 75), alpha = 0.75) +
  theme_classic() +
  labs(title = "Simulated negative binomial distribution", fill = "Mean", y = "Frequency", x = "Mean")

@@

Now let's first plot the distribution of unfiltered reads from our dataset:

<br>

@@

The table of read counts above has rows representing individual transcripts and columns representing different replicate samples across treatments.

<br>

```{r, message = FALSE, warning = FALSE}
# Name experimental samples: names correspond to pCO2 treatment (300, 600, 900) + days (12 h2 vs 7 days) + replicate ID
colnames(data) <- c("300.7.a", "300.7.b", "300.7.c","300.12.a", "300.12.b", "300.12.c", 
                    "900.7.a", "900.7.b", "900.7.c", "900.12.a", "900.12.b", "900.12.c", 
                    "600.7.a", "600.7.b", "600.7.c", "600.12.a", "600.12.b", "600.12.c")

## Create 'targets' and 'Group dataframe, expressing experimental variables for DEG analysis
pCO2 <- as.numeric(c( 255, 255, 255, 255, 255, 255,
                      530, 530, 530, 530, 530, 530,
                      918, 918, 918, 918, 918, 918))

treatment <- c("B","B", "B", "B", "B", "B",
               "R", "R", "R", "R", "R", "R",
               "Y", "Y", "Y", "Y", "Y", "Y")

day <- as.numeric(c(7, 7, 7, .5, .5, .5,
                    7, 7, 7, .5, .5, .5,
                    7, 7, 7, .5, .5, .5))

targets <- data.frame(pCO2, day, treatment)
targets$grouping <- paste(targets$pCO2, targets$day, sep = ".")

# The group factor represents the combined levels of an experimental replicate across all variables
Group <- factor(paste(targets$day, targets$pCO2, sep = "_"))

# Data must be rounded to nearest integer in order to be fit for negative binomial distribution
data_input <- round(data)

# Peak at rounded data
head(data_input)

# Plot distribution of unfiltered read counts across all samples 
ggplot(data = data.frame(rowMeans(data_input)),
       aes(x = rowMeans.data_input.)) +
  geom_histogram(fill = "grey") +
  xlim(0, 500) +
  theme_classic() +
  labs(title = "Distribution of unfiltered reads") +
  labs(y = "Density", x = "Raw read counts",
  title = "Read count distribution: untransformed, unnormalized, unfiltered")

@@

As you can see in the above plot, the raw distribution of all read counts takes on a left-skewed negative binomial distribution similar to the purple distribution in the example above. While looking at the distribution of your raw reads is useful, these are not in fact the data you will be inputting to tests of differential expression. Let's plot the distribution of filtered reads normalized by library size, expressed as log2 counts per million reads (logCPM). These are the reads we will use in our test, and after we plot their distribution, we will conduct one more, slightly more robust, test of our data's fit to the negative binomial distribution using residuals from fitted models.

<br>

@@

#Let's remove samples with less than 0.5 cpm (this is ~10 counts in the count file) in fewer then 9/12 samples
keep <- rowSums(cpm(y) > .5) >= 9

table(keep)

# Set keep.lib.sizes = F and recalculate library sizes after filtering
y <- y[keep, keep.lib.sizes = FALSE]

y <- calcNormFactors(y)

# Calculate logCPM
df_log <- cpm(y, log = TRUE, prior.count = 2)

# Plot distribution of filtered logCPM values
ggplot(data = data.frame(rowMeans(df_log)), 
       aes(x = rowMeans.df_log.) ) +
  geom_histogram(fill = "grey") +
  theme_classic() +
  labs(y = "Density", x = "Filtered read counts (logCPM)",
       title = "Distribution of normalized, filtered read counts")

@@

Our raw data appear to follow a strong negative binomial distribution! Later, we will see whether this holds for residuals from fitted negative binomial GLMs.

<br>

@@

Before analyzing our data, it is essential that we look at the multivariate relationships between our samples based on transcriptome-wide expression levels. Below is example code and output for a principal coordinates analysis (PCOA) plot that visualizes multifactorial RNA-seq replicates according to two predictor variables across major and minor latent variables or PCOA axes. These predictor variables, as discussed above, are *p*CO${_2}$ and time-in-treatment.

<br>

@@

# Export pcoa loadings
dds.pcoa = pcoa(vegdist(t(df_log <- cpm(y, log = TRUE, prior.count = 2)),
                          method = "euclidean") / 1000)

@@

# Calculate % variation explained by each eigenvector
percent <- dds.pcoa$values$Eigenvalues
cumulative_percent_variance <- (percent / sum( percent)) * 100

# Prepare information for pcoa plot, then plot
color <- c("steelblue1", "tomato1", "goldenrod1")
par(mfrow = c(1, 1))
plot(
  scores[, 1],
  scores[, 2],
  cex = .5,
  cex.axis = 1,
  cex.lab = 1.25,
  xlab = paste("PC1, ", round(cumulative_percent_variance[1], 2), "%"),
  ylab = paste("PC2, ", round(cumulative_percent_variance[2], 2), "%")
  )

# Add visual groupings to pcoa plot
ordihull(
  scores,
  as.factor(targets$treatment),
  border = NULL,
  lty = 2,
  lwd = .5,
  label = F,
  col = color,
  draw = "polygon",
  alpha = 100,
  cex = .5
  )

ordispider(scores, as.factor(targets$grouping), label = F) # Vectors connecting samples in same pCO2 x time group

ordilabel(scores, cex = 0.5) # Label sample IDs

@@

In the PCOA above, blue = low *p*$CO_{2}$, yellow = moderate *p*$CO_{2}$, and red = high *p*$CO_{2}$. From this plot, we can see that treatment and time both influence multivariate gene expression across the RNA-seq samples. By and large, samples cluster according to these two predictors in a manner consistent with what we would expect from our experimental design if *p*$CO_{2}$ and time were to affect gene expression.

<br>

@@

logCPM.pca <- prcomp(t (df_log))
logCPM.pca.proportionvariances <-
((logCPM.pca$sdev ^ 2) / (sum(logCPM.pca$sdev ^ 2))) * 100

## Do treatment groups fully segregate? Wrap samples by pCO2 x time, not just pCO2
# Replot using logCPM.pca
plot(
  logCPM.pca$x,
  type = "n",
  main = NA,
  xlab = paste("PC1, ", round(logCPM.pca.proportionvariances[1], 2), "%"),
  ylab = paste("PC2, ", round(logCPM.pca.proportionvariances[2], 2), "%")
  )

points(logCPM.pca$x,
       col = "black",
       pch = 16,
       cex = 1)
       colors2 <-
       c("steelblue1",
         "dodgerblue2",
         "tomato1",
         "coral",
         "goldenrod1",
         "goldenrod3")

       ordihull(
       logCPM.pca$x,
       targets$grouping,
       border = NULL,
       lty = 2,
       lwd = .5,
       col = colors2,
       draw = "polygon",
       alpha = 75,
       cex = .5,
       label = T
       )

@@

This plot is only a slight adjustment of the first PCOA. If you'd like to wrap each multifactorial group and get a better idea of potential overlap between clusters, the plot and code above can be applied to do so.

Now that we have conducted QC analyses and plots of the read count data, it is time to begin model fitting and testing for differential expression.

<br>

@@

Gene expression traits can exhibit non-linear performance curves across a continuous variable much like other physiological traits. Studies that incorporate more than two levels of a continuous predictor provide an opportunity to test for non-linear effects on gene expression. Similar to performance traits such as growth or metabolic rates, gene expression can be modelled across a continuous variable by fitting non-linear 'performance curves' to read count data. Rivera *et al.* 2021 recently discussed the benefits of fitting non-linear reaction norms to gene expression data in order to better model acclimation by tolerant and sensitive genotypes to an environmental stressor. In a conceptual figure of theirs, shown below, they depict non-linear variation in expression before, during, and after exposure to a stressor for four reaction norm phenotypes: (i) a tolerant conspecific that frontloads (e.g., constitutively upregulates) stress-induced transcripts, (ii) a tolerant conspecific that exhibits reduced overall expression of stress-responsive transcripts, (iii) a sensitive conspecific that upregulates inducible transcripts and cannot recover post-stress, and (iv) a resilient conspecific that mounts a transcriptional stress response followed by recovery and downregulation [@Rivera2021-hn].

<br>

```{r, echo = FALSE, out.height = "40%", fig.align = "center", }

knitr::include_graphics("images/Rivera_etal_fig.png")

```
From Rivera *et al.* 2021 - "Transcriptomic resilience and timing. (a) Gene expression reaction norms of four strategies during recovery after a stressor. We use triangles again for patterns that may confer tolerance and circles for patterns associated with stress sensitivity. While all triangle paths show a return to baseline (resilience) the pink (frontloading) and yellow (dampening) are also depicting differences in baseline and plasticity and are therefore labelled differently. (b) Adapted from the rolling ball analogy commonly used for ecological resilience and depicted in Hodgson et al. (2015). Each ball represents a gene showing a color-matched expression pattern in (a). Landscapes represent expression possibilities during a stress event. In the absence of stress, the ball will settle in a trough, representing baseline expression levels. Elasticity (rate of return to the baseline) is represented by the size of the arrow (i.e., larger arrows have faster rates of return). Pink dotted line is the expression landscape for the frontloaded ball. (c) Using Torres et al. (2016) loops through disease space as an alternative framework of an organism's path through stress response and recovery. The color gradient represents the resulting phenotype for a given path through stress and recovery space, though x-and y-axis can denote any two parameters that are correlated but with a time lag."

Non-linear reaction norms can be modelled across time, such as Rivera *et al.* have presented, or across multiple levels of an abiotic variable such as temperature, pH, etc. in order to fit a performance curve. Performance curves are a fundamental tool in ecological physiology, and enable more robust hypothesis testing in RNA-seq studies of environmental acclimation. For example, testing for variation in the shape of gene expression performance curves between early and long-term exposure timepoints can provide critical information about the role of gene expression in acclimation. Below we outline how to fit and test for non-linear gene expression performance curves in multifactorial RNA-seq experiments using examples in edgeR, DESeq2, and custom code.

One of the simplest non-linear relationships that can be fitted to the expression of a transcript across an continuous variable is a second-order polynomial, otherwise known as a quadratic function, which can be expressed as $y_{i} = \mu + \beta_{1}x^2 + \beta_{2}x$ where $y$ is the abundance of a given transcript ($i$), $\mu$ is the intercept, and $y$ is the continuous variable. For the parabola generated by fitting a second-order polynomial, $\beta_{1}$ > 0 opens the parabola upwards while $\beta_{1}$ < 0 opens the parabola downwards. The vertex of the parabola is controlled by $\beta_{2}$ such that when $\beta_{1}$ is negative, greater $\beta_{2}$ values result in the vertex falling at higher values of $x$.

Quadratic polynomials applied to phenotypic performance curves commonly possess negative $\beta_{1}$ values with positive $\beta_{2}$ values: a downard-opening parabola with a positive vertex. However, quadtratic curves fitted to gene expression data can take on a variety of postiive or negative forms similar to exponential curves, saturating curves, and parabolas. For instance, the expression of a gene may peak an intermediate level of an environmental level before crashing or it may exponentially decline across that variable. Such trends may better model changes in the transcription of a gene compared to a linear model. To get started, we will fit non-linear second order polynomials before testing for whether model predictions for a given gene are significantly improved by a non-linear linear model.

<br>

@@

Let's fit a second-order polynomial for the effect of *p*CO$_{2}$ using edgeR. Using differential expression tests, we will then determine whether *p*CO$_{2}$ affected a gene's rate of change in expression and expression maximum by applying differential expression tests to $\beta_{1}$ and $\beta_{2}$ parameters. By testing for differential expression attributed to intereactions between time and $\beta_{1}$ or $\beta_{2}$, we will then test for whether these parameters were significantly different across exposure times such that 0.5 days and 7 days of acclimation to *p*CO$_{2}$ altered the rate of change in expression across *p*CO$_{2}$ ($\beta_{1}$) or the maximum of expression ($\beta_{2}$).

Since this is the first time we're fitting models to our data in this walkthrough, we will also output some important plots to help us explore and QC model predictions as we work toward testing for differential expression attributed to non-linear effects.

<br>

@@

# Square pCO2 variable
pCO2_2 <- pCO2 ^ 2

# Estimate dispersion coefficients
y1 <- estimateDisp(y, robust = TRUE) # Estimate mean dispersal

# Plot tagwise dispersal and impose w/ mean dispersal and trendline
plotBCV(y1)

@@

Our residuals appear to be normally distributed, indicating that our data fit the negative binomial distribution assumed by the GLM.

Now let's get back to analyzing non-linear effects on gene expression!

<br>

@@
